---
title: Manipulating tibbles with dplyr
teaching: 40
exercises: 15
questions:
- "How can I manipulate tibbles without repeating myself?"
objectives:
- " To be able to use the six main `dplyr` data manipulation 'verbs' with pipes."
keypoints:
- "Use the `dplyr` package to manipulate tibbles."
- "Use `select()` to choose variables from a tibbles."
- "Use `filter()` to choose data based on values."
- "Use `group_by()` and `summarize()` to work with subsets of data."
- "Use `mutate()` to create new variables."
source: Rmd
---

```{r, echo=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
```

In the previous episode we used the `readr` package to load tabular data into a tibble within R.  The `readr` package is part of a family of packages known as the   [tidyverse](http://tidyverse.org/).  The tidyverse packages are designed to work well together; they provide a modern and streamlined approach to data-analysis, and deal with some of the idiosyncrasies of base R.

We load the core libraries in the tidyverse with:

```{r}
library("tidyverse")
```

This loads the most commonly used packages in the tidyverse; we used `readr` in the previous episode.  We will cover all of the other main packages, with the exception of `purrr` in this course. There are other [libraries included](https://github.com/tidyverse/tidyverse) but these are less widely used, and must be loaded manually if they are required; these aren't covered in this course. 

Let's dive in and look at how we can use the tidyverse to analyse and, in a couple of episodes' time,  plot data from the [gapminder project](https://www.gapminder.org/).  At [the start of the course]({{ page.root}}/02-project-intro), you should have copied the file `gapminder-FiveYearData.csv` to your `data` directory.     Take a look at it using a text editor such as notepad.   The first line contains variable names, and values are separated by commas.  Each record starts on a new line. 

As with the [lesson on data structures]({{ page.root }}/04-data-structures-part1) we use the `read_csv()` function to load data from a comma separated file:

```{r}
gapminder <- read_csv("./data/gapminder-FiveYearData.csv")
```

As we discussed in [lesson 4]({{ page.root }}/04-data-structures-part1/), variables in R can be character, integer, double, etc.   A tibble (and R's built in equivalent; the data-frame) require that all the values in a particular column have the same data type.  The `read_csv()` function will attempt to infer the data type of each column, and prints the column types it has guessed to the screen.  If the wrong column types have been generated, you can pass the `col_types=` option to `read_csv()`.  

For example, if we wanted to load the `pop` column as a character string, we would use:

```{r}
gapminderPopChar <- read_csv("./data/gapminder-FiveYearData.csv", 
                             col_types = cols(
                               country = col_character(),
                               year = col_integer(),
                               pop = col_character(),
                               continent = col_character(),
                               lifeExp = col_double(),
                               gdpPercap = col_double()
) )
```

> ## Setting column types
> 
> Try reading a file using the `read_csv()` defaults (i.e. guessing column types).
> If this fails you can cut and paste the guessed column specification, and modify
> this with the correct column types.  It is good practice to do this anyway; it makes
> the data types of your columns explicit, and will help protect you if the format 
> of your data changes.
{: .callout}

## Manipulating tibbles 

Manipulation of tibbles means many things to many researchers. We often
select only certain observations (rows) or variables (columns). We often group the
data by a certain variable(s), or calculate summary statistics.

## The `dplyr` package

The  [`dplyr`](https://cran.r-project.org/web/packages/dplyr/dplyr.pdf)
package is part of the tidyverse.  It provides a number of very useful functions for manipulating tibbles (and their base-R cousin, the `data.frame`) 
in a way that will reduce repetition, reduce the probability of making
errors, and probably even save you some typing. 

We will cover:

1. selecting variables with `select()`
2. subsetting observations with `filter()`
3. grouping observations with `group_by()`
4. generating summary statistics using `summarize()`
5. generating new variables using `mutate()`
6. Sorting tibbles using `arrange()`
7. chaining operations together using pipes `%>%` 

## Using `select()`

If, for example, we wanted to move forward with only a few of the variables in
our tibble we use the `select()` function. This will keep only the
variables you select.

```{r}
year_country_gdp <- select(gapminder,year,country,gdpPercap)
print(year_country_gdp)
```

Select will select _columns_ of data.  What if we want to select rows that meet certain criteria?  

## Using `filter()`

The `filter()` function is used to select rows of data.  For example, to select only countries in 
Europe:

```{r}
gapminder_Europe <- filter(gapminder, continent=="Europe") 
print(gapminder_Europe)
```

Only rows of the data where the condition (i.e. `continent=="Europe") is `TRUE` are kept.

## Using pipes and dplyr

We've now seen how to choose certain columns of data (using `select()`) and certain rows of data (using `filter()`).  In an analysis we often want to do both of these things (and many other things, like calculating summary statistics, which we'll come to shortly).    How do we combine these?

There are several ways of doing this; the method we will lean about today is using _pipes_.  

The pipe operator `%>%` lets us pipe the output of one command into the next.   This allows us to build up a data-processing pipeline.  This approach has several advantages:

* We can build the pipeline piecemeal - building the pipeline step-by-step is easier than trying to 
perform a complex series of operations in one go
* It is easy to modify and reuse the pipeline
* We don't have to make temporary tibbles as the analysis progresses

> ## Pipelines and the shell
>
> If you're familiar with the Unix shell, you may already have used pipes to
> pass the output from one command to the next.  The concept is the same, except
> the shell uses the `|` character rather than R's pipe operator `%>%`
{: .callout}


> ## Keyboard shortcuts and getting help
> 
> The pipe operator can be tedious to type.  In Rstudio pressing <kbd>Ctrl</kbd> + <kbd>Shift</kbd>+<kbd>M</kbd> under
> Windows / Linux will insert the pipe operator.  On the mac, use <kbd>&#8984;</kbd> + <kbd>Shift</kbd>+<kbd>M</kbd>.
>
> We can use tab completion to complete variable names when entering commands.
> This saves typing and reduces the risk of error.
> 
> RStudio includes a helpful "cheat sheet", which summarises the main functionality
> and syntax of `dplyr` and the related package `tidyr`.  This can be accessed via the
> help menu --> cheatsheets --> data manipulation with dplyr, tidyr. 
> (In RStudio 1.1 this has been replaced with the cheat sheet "data transformation with dplyr".)
>
{: .callout}

Let's rewrite the select command example using the pipe operator:

```{r}
year_country_gdp <- gapminder %>% select(year,country,gdpPercap)
print(year_country_gdp)
```

To help you understand why we wrote that in that way, let's walk through it step
by step. First we summon the gapminder tibble and pass it on, using the pipe
symbol `%>%`, to the next step, which is the `select()` function. In this case
we don't specify which data object we use in the `select()` function since in
gets that from the previous pipe. 

What if we wanted to combine this with the filter example? I.e. we want to select year, country and GDP per capita, but only for countries in Europe?  We can join these two operations using a pipe; feeding the output of one command directly into the next:


```{r}
year_country_gdp_euro <- gapminder %>%
    filter(continent == "Europe") %>%
    select(year,country,gdpPercap)
print(year_country_gdp_euro)
```

Note that the order of these operations matters; if we reversed the order of the `select()` and `filter()` functions, the `continent` variable wouldn't exist in the data-set when we came to apply the filter.  

What about if we wanted to match more than one item?  To do this we use the `%in%` operator:

```{r}
gapminder_benelux <- gapminder %>% 
  filter(country %in% c("Belgium", "Netherlands", "France"))
```


> ## Another way of thinking about pipes
>
> It might be useful to think of the statement
> ```{r, eval=FALSE}
>  gapminder %>%
>     filter(continent=="Europe") %>%
>     select(year,country,gdpPercap)
> ```
>  as a sentence, which we can read as
> "take the gapminder data *and then* `filter` records where continent == Europe
> *and then* `select` the year, country and gdpPercap
> 
> We can think of the `filter()` and `select()` functions as verbs in the sentence; 
> they do things to the data flowing through the pipeline.  
>
{: .callout}

> ## Splitting your commands over multiple lines
> 
> It's generally a good idea to put one command per line when
> writing your analyses.  This makes them easier to read.   When
> doing this, it's important that the `%>%` goes at the _end_ of the
> line, as in the example above.  If we put it at the beginning of a line, e.g.:
> 
> ```{r}
> gapminder_benelux <- gapminder 
> %>% filter(country %in% c("Belgium", "Netherlands", "France"))
> ```
> 
> the first line makes a valid R command.  R will then treat the next line 
> as a new command, which won't work.
{: .callout}


> ## Challenge 1
>
> Write a single command (which can span multiple lines and includes pipes) that
> will produce a tibble that has the values of `lifeExp`, `country`
> and `year`, for the countries in Africa, but not for other Continents.  How many rows does your tibble  
> have?
>
> > ## Solution to Challenge 1
> >```{r}
> >year_country_lifeExp_Africa <- gapminder %>%
> >                            filter(continent=="Africa") %>%
> >                            select(year,country,lifeExp)
> > nrow(year_country_lifeExp_Africa)
> > ```
> > As with last time, first we pass the gapminder tibble to the `filter()`
> > function, then we pass the filtered version of the gapminder tibble  to the
> > `select()` function. **Note:** The order of operations is very important in this
> > case. If we used 'select' first, filter would not be able to find the variable
> > continent since we would have removed it in the previous step.
> {: .solution}
{: .challenge}


## Sorting tibbles

The `arrange()` function will sort a tibble by one or more of the variables in it:

```{r}
gapminder %>%
  filter(continent == "Europe", year == 2007) %>% 
  arrange(pop)
```
We can use the `desc()` function to sort a variable in reverse order:

```{r}
gapminder %>%
  filter(continent == "Europe", year == 2007) %>% 
  arrange(desc(pop))
```

## Generating new variables

The `mutate()` function lets us add new variables to our tibble.  It will often be the case that these are variables we _derive_ from existing variables in the data-frame. 

As an example, the gapminder data contains the population of each country, and its GDP per capita.  We can use this to calculate the total GDP of each country:

```{r}
gapminder_totalgdp <- gapminder %>% 
  mutate(gdp = gdpPercap * pop)

```

> ## Challenge 2
> 
> Create a tibble containing each country in Europe, its life expectancy in 2007
> and the rank of the country's life expectancy. 
>
> Hint: First filter and then use mutate to create a new variable with the rank in it.  The cheat-sheet contains useful
> functions you can use when you make new variables (the cheat-sheets can be found in the help menu in RStudio)
>
> Can you reverse the ranking order
> so that the country with the longest life expectancy gets the lowest rank?
> Hint: This is similar to sorting in reverse order
>
> > ## Solution to challenge 2
> > ```{r}
> > europeLifeExp <- gapminder %>% 
> >   filter(continent == "Europe", year == 2007) %>% 
> >   select(country, lifeExp) %>% 
> >   mutate(rank = min_rank(lifeExp))
> > print(europeLifeExp, n=100)
> > ```
> > 
> > To reverse the order of the ranking, use the `desc` function, i.e.
> > `mutate(rank = min_rank(desc(lifeExp)))`
> > 
> > There are several functions for calculating ranks; you may have used, e.g. `dense_rank()`
> > The functions handle ties differently.  The help file for `dplyr`'s ranking functions
> > explains the differences, and can be accessed with `?ranking`
> {: .solution}
{: .challenge}

## Calculating summary statistics

We often wish to calculate a summary statistic (the mean, standard deviation, etc.)
for a variable.  We frequently want to calculate a separate summary statistic for several
groups of data (e.g. the experiment and control group).    We can calculate a summary statistic
for the whole data-set using the dplyr's `summarise()` function:

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  summarise(meanlife = mean(lifeExp), medianlife = median(lifeExp))
```

To generate summary statistics for each value of another variable we use the 
`group_by()` function:

```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarise(meanlife = mean(lifeExp), medianlife = median(lifeExp))
```

> ## Challenge 3
>
> For each combination of continent and year, calculate the average life expectancy.
>
> > ## Solution to Challenge 3
> >
> >```{r}
> > lifeExp_bycontinentyear <- gapminder %>% 
> >    group_by(continent, year) %>% 
> >   summarise(mean_lifeExp = mean(lifeExp))
> > print(lifeExp_bycontinentyear)
> >```
> {: .solution}
{: .challenge}

## `count()` and `n()`
A very common operation is to count the number of observations for each
group. The `dplyr` package comes with two related functions that help with this.


If we need to use the number of observations in calculations, the `n()` function
is useful. For instance, if we wanted to get the standard error of the life
expectancy per continent:

```{r}
gapminder %>%
    filter(year == 2002) %>%	
    group_by(continent) %>%
    summarize(se_pop = sd(lifeExp)/sqrt(n()))
```

Although we could use the `group_by()`, `n()` and `summarize()` functions to calculate the number of observations in each group, `dplyr` provides the `count()` function which automatically groups the data, calculate the totals and then ungroups it. 

For instance, if we wanted to check the number of countries included in the
dataset for the year 2002, we can use:

```{r}
gapminder %>%
    filter(year == 2002) %>%
    count(continent, sort = TRUE)
```
We can optionally sort the results in descending order by adding `sort=TRUE`:



## Connect mutate with logical filtering: `ifelse()`

When creating new variables, we can hook this with a logical condition. A simple combination of 
`mutate()` and `ifelse()` facilitates filtering right where it is needed: in the moment of creating something new.
This easy-to-read statement is a fast and powerful way of discarding certain data (even though the overall dimension
of the tibble will not change) or for updating values depending on this given condition.

```{r}
## keeping all data but "filtering" after a certain condition
# calculate GDP only for people with a life expectation above 25
gdp_pop_bycontinents_byyear_above25 <- gapminder %>%
    mutate(gdp_billion = ifelse(lifeExp > 25, gdpPercap * pop / 10^9, NA)) %>%
    group_by(continent, year) %>%
    summarize(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop),
              mean_gdp_billion = mean(gdp_billion),
              sd_gdp_billion = sd(gdp_billion))

## updating only if certain condition is fullfilled
# for life expectations above 40 years, the gpd to be expected in the future is scaled
gdp_future_bycontinents_byyear_high_lifeExp <- gapminder %>%
    mutate(gdp_futureExpectation = ifelse(lifeExp > 40, gdpPercap * 1.5, gdpPercap)) %>%
    group_by(continent, year) %>%
    summarize(mean_gdpPercap = mean(gdpPercap),
              mean_gdpPercap_expected = mean(gdp_futureExpectation))
```


> ## Challenge 4
>
> Select two countries at random from each continent.  For each continent, calculate the 
> average life expectancy of the two countries.   Then arrange the continent names in reverse order.
> 
> Hint: The `dplyr` function `sample_n()` can be used to select a sample of rows.  
>
> The final tibble should contain the columns `continent` and the average life expectancy. You do
> *not* need to include the countries that were selected in it.
>
> > ## Solution to Challenge 4
> >```{r}
> >set.seed(171124)
> >lifeExp_2countries_bycontinents <- gapminder %>%
> >    filter(year==2002) %>%
> >    group_by(continent) %>%
> >    sample_n(2) %>%
> >    summarize(mean_lifeExp=mean(lifeExp)) %>%
> >    arrange(desc(mean_lifeExp))
> >lifeExp_2countries_bycontinents 
> >```
> >
> > Discussion: Do get the same answer as your neighbour?  What about if you run the command again? Do you get the
> > same answer as last time?
> > 
> > As we're sampling the rows at random we expect to get a different answer from our neighbour, and each time
> > we run the command.   You can set the random number _seed_ used by R using `set.seed(**a number**)`, as shown in the example above.
> > By default
> > R generates a seed when one is first requried using the process ID of R and the current time (i.e. essentially at
> > random). If you are using random numbers in your work you should set the seed at the start of your analysis
> > so that your results are reproducible. 
> {: .solution}
{: .challenge}


> ## Equivalent functions in base R
>
> In this course we've taught the tidyverse.  You are likely come across
> code written others in base R.  You can find a guide to some base R functions
> and their tidyverse equivalents [here](http://www.significantdigits.org/2017/10/switching-from-base-r-to-tidyverse/),
> which may be useful when reading their code.
>
{: .callout}
## Other great resources

* [R for Data Science](http://r4ds.had.co.nz/)
* [Data Wrangling Cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
* [Introduction to dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html) - this is the package vignette.  It can be viewed within R using `vignette(package="dplyr", "dplyr")`
* [Data wrangling with R and RStudio](https://www.rstudio.com/resources/webinars/data-wrangling-with-r-and-rstudio/)
