---
title: Functions Explained
teaching: 45
exercises: 15
questions:
- "How can I write a new function in R?"
objectives:
- "Define a function that takes arguments."
- "Return a value from a function."
- "Test a function."
- "Set default values for function arguments."
- "Explain why we should divide programs into small, single-purpose functions."
keypoints:
- "Use `function` to define a new function in R."
- "Use parameters to pass values into functions."
- "Load functions into programs using `source`."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
library(tidyverse)
knitr_fig_path("07-")
# Silently load in the data so the rest of the lesson works
gapminder <- read_csv("data/gapminder-FiveYearData.csv")
```

If we only had one data set to analyse, it would probably be faster to load the
file into a spreadsheet and use that to plot simple statistics. However, the
gapminder data is updated periodically, and we may want to pull in that new
information later and re-run our analysis again. We may also obtain similar data
from a different source in the future.

In this lesson, we'll learn how to write a function so that we can repeat
several operations with a single command.

> ## What is a function?
>
> Functions gather a sequence of operations into a whole, preserving it for ongoing use. Functions provide:
>
> * a name we can remember and invoke it by
> * relief from the need to remember the individual operations
> * a defined set of inputs and expected outputs
> * rich connections to the larger programming environment
>
> As the basic building block of most programming languages, user-defined
> functions constitute "programming" as much as any single abstraction can. If
> you have written a function, you are a computer programmer.
{: .callout}

## Defining a function

Let's create a new R script file in the `src/` directory and call it functions-lesson.R:

```{r}
my_sum <- function(a, b) {
  the_sum <- a + b
  return(the_sum)
}
```

We can use the function in the same way as any other functions we've used so far in this course:

```{r}
my_sum(2,3)
my_sum(10,10)
```


Now let’s define a function `fahr_to_kelvin` that converts temperatures from Fahrenheit to Kelvin:

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

We define `fahr_to_kelvin` by assigning it to the output of `function`.  The
list of argument names are contained within parentheses.  Next, the
[body]({{ page.root }}/reference/#function-body) of the function--the statements that are
executed when it runs--is contained within curly braces (`{}`).  The statements
in the body are indented by two spaces.  This makes the code easier to read but
does not affect how the code operates.

When we call the function, the values we pass to it as arguments are assigned to those
variables so that we can use them inside the function.  Inside the function, we
use a [return statement]({{ page.root }}/reference/#return-statement) to send a result back
to whoever asked for it.

> ## Tip
>
> One feature unique to R is that the return statement is not required.
> R automatically returns whichever variable is on the last line of the body
> of the function. But for clarity, we will explicitly define the
> return statement.
{: .callout}


Let's try running our function.
Calling our own function is no different from calling any other function:

```{r}
# freezing point of water
fahr_to_kelvin(32)
```

```{r}
# boiling point of water
fahr_to_kelvin(212)
```

> ## Challenge 1
>
> Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin
> and returns that temperature in Celsius
>
> Hint: To convert from Kelvin to Celsius you subtract 273.15
>
> > ## Solution to challenge 1
> >
> > Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin
> > and returns that temperature in Celsius
> >
> > ```{r}
> > kelvin_to_celsius <- function(temp) {
> >  celsius <- temp - 273.15
> >  return(celsius)
> > }
> > ```
> {: .solution}
{: .challenge}

## Combining functions

The real power of functions comes from mixing, matching and combining them
into ever-larger chunks to get the effect we want.

> ## Challenge 2
>
> Define the function to convert directly from Fahrenheit to Celsius,
> by using the `fahr_to_kelvin` function defined earlier, and the 
> `kelvin_to_celsius` function you wrote in challenge 1. 
>
> > ## Solution to challenge 2
> >
> > Define the function to convert directly from Fahrenheit to Celsius,
> > by reusing these two functions above
> >
> >
> > ```{r}
> > fahr_to_celsius <- function(temp) {
> >   temp_k <- fahr_to_kelvin(temp)
> >   result <- kelvin_to_celsius(temp_k)
> >   return(result)
> > }
> > ```
> {: .solution}
{: .challenge}


## Programming with dplyr

Let's write a function that will filter our gapminder data by year *and* calculate the GDP of all the countries 
that are returned.   First, let's think about how we'd do this *without* using a function.  We'll then put our 
code in a function; this will let us easily repeat the calculation for other years, without re-writing the code.

This saves us effort, but more importantly it means that the code for calculating the GDP only exists in a single place.  If we find we've used the wrong formula to calculate GDP we would only need to change it in a single place, rather than hunting down all the places we've calculated it.  This important idea is known as "Don't Repeat Yourself".

We can filter and calculate the GDP as follows:

```{r}
gdpgapfiltered <- gapminder %>%
    filter(year == 2007) %>% 
    mutate(gdp = gdpPercap * pop)

```

Note that we filter by year first, and _then_ calculate the GDP.  Although we could reverse the order of the `filter()` and `mutate()` functions, it is more efficient to filter the data and then calculate the GDP on the remaining data.

Let's put this code in a function:

```{r}
calc_GDP_and_filter <- function(dat, year){
  
  gdpgapfiltered <- dat %>%
      filter(year == year) %>% 
      mutate(gdp = gdpPercap * pop)
  
  return(gdpgapfiltered)
  
}

calc_GDP_and_filter(gapminder, 1997)

```

The function hasn't done what we'd expect; we've successfully passed in the gapminder data, via the `dat` parameter,
but it looks like the `year` parameter has been ignored.

Look at the line: 

```{r, eval=FALSE}
    filter(year == year) %>% 
```

We passed a value of `year` into the function when we called it.  R has no way of knowing we're referring to the function parameter's `year` rather than the tibble's `year` variable.     We run into this problem because `dplyr` uses what's known as "Non Standard Evaluation" (NSE); this means that the package doesn't follow R's usual rules about how parameters are evaluated.  

NSE is usually really useful; when we've written things like `gapminder %>% select(year, country)` we've made use of non standard evaluation.  This is much more intuitive than the base R equivalent, where we'd have to write something like `gapminder[, names(gapminder) %in% c("year", "country")]`.  Unfortunately this simplicity comes at a price when we come to write functions.  It means we need a way of telling R whether we're referring to a variable in the tibble, or a parameter we've passed via the function.

We can use the `calc_GDP_and_filter`'s `year` parameter like a normal variable in our function _except_ when we're using it as part of a parameter to a `dplyr` verb (e.g. `filter`).   We need to _unquote_ the `year` parameter so that the `dplyr` function can see its contents (i.e. 1997 in this example).  We do this using the `UQ()` ("unquote") function:

```{r, eval=FALSE}
    filter(year == UQ(year) ) %>% 
```

When the filter function is evaluated it will see:


```{r, eval=FALSE}
    filter(year == 1997) %>% 
```


```{r}
calc_GDP_and_filter <- function(dat, year){
  
gdpgapfiltered <- dat %>%
    filter(year == UQ(year) ) %>% 
    mutate(gdp = gdpPercap * pop)

return(gdpgapfiltered)
  
}

calc_GDP_and_filter(gapminder, 1997)
                 

```

Our function now works as expected.

> ## Another way of thinking about quoting
> 
> (This section is based on [programming with dplyr](http://dplyr.tidyverse.org/articles/programming.html))
>
> Consider the following code:
>
> ```{r}
> greet <- function(person){
>   print("Hello person")
> }
> 
> greet("David")
> ```
> 
> The `print` function has no way of knowing that `person` refers to the variable `person`, and isn't 
> part of the string `person`.  To make the contents of the `person` variable visible to the function we
> need to construct the string, using the `paste` function:
>
> ```{r}
> greet <- function(person){
>   print(paste("Hello", person))
> }
> 
> greet("David")
> ```
> This means that the `person` variable is evaluated in an _unquoted_ environment, so its contents can be evaluated.
{: .callout}

There is one small issue that remains; how does filter _know_ that the first `year` in ``` filter(year == UQ(year) ) %>%  ``` refers to the `year` variable in the tibble?  What happens if we delete the 
year variable? Surely this should give an error?

```{r}
gap_noyear <- gapminder %>% select(-year)
calc_GDP_and_filter(gap_noyear, 1997)
```

As you can see, it doesn't; instead the `filter()` function will "fall through" to look for the `year` variable in `filter()`'s _calling environment_.  This is the `calc_GDP_and_filter()` environment, which does have a `year` variable.  Since this is a standard R variable, it will be implicitly unquoted, so `filter()` will see:
```{r, eval=FALSE}
    filter(1997 == 1997) %>% 
```
which is `TRUE`, so the filter won't do anything!

We need a way of telling the function that the first `year` "belongs" to the data.  We can do this with the `.data` pronoun:

```{r}
calc_GDP_and_filter <- function(dat, year){
  
gdpgapfiltered <- dat %>%
    filter(.data$year == UQ(year) ) %>% 
    mutate(gdp = .data$gdpPercap * .data$pop)

return(gdpgapfiltered)
  
}

calc_GDP_and_filter(gapminder, 1997)
calc_GDP_and_filter(gap_noyear, 1997)

```                 


As you can see, we've also used the `.data` pronoun when calculating the GDP; if our tibble was missing either the `gdpPercap` or `pop` variables, R would search in the calling environment (i.e. the `calc_GDP_and_filter()` function).  As the variables aren't found there it would look in the `calc_GDP_and_filter()`'s calling environment, and so on.  If it finds variables matching these names, they would be used instead, giving an incorrect result; if they cannot be found we will get an error.  Using the `.data` pronoun makes our source of the data clear, and prevents this risk.


> ## Challenge:  Filtering by country name and year
>
> Create a new function that will filter by country name *and* by year, and then calculate the GDP
> You can assume that both name and year will be provided; we will cover how to check this is true
> in the next section
>
> > ## Solution
> >
> > ```{r}
> >  calcGDPCountryYearFilter <- function(dat, year, country){
> >  dat <- dat %>% filter(.data$year == UQ(year) ) %>% 
> >        filter(.data$country == UQ(country) ) %>% 
> >         mutate(gdp = .data$pop * .data$gdpPercap)
> >         
> >  return(dat)
> > }
> > calcGDPCountryYearFilter(gapminder, year=2007, country="United Kingdom")
> > ```
> > 
> {: .solution}
{: .challenge}


## Making our function more flexible

At the moment, the function we wrote in the challenge will calculate the GDP for a single country and year.  It would be more flexible and useful if we could calculate the GDP for:

 * The whole dataset; (if we don't provide any years or countries)
 * All the countries, filtered by one or more years; (if we only provide years)
 * All the years, filtered by one or more countries; (if we only provide countries)
 * A combination of one or more years and one or more countries. (if we provide both)

> ## Discussion
>
> What happens if we try calling the `calcGDPCountryYearFilter()` function without `year` or `country` parameters?
>
> > ## Solution
> > 
> > ```{r}
> > calcGDPCountryYearFilter(gapminder)
> >```
> >
> {: .solution}
{: .discussion}

We can pass a default parameter for year and country; this means that if we call the function without specifying the parameter the default will be used instead.  By setting the defaults to `NULL`, and then testing whether each parameter is `NULL` we can call the appropriate parts of the function to filter by `year` and / or `country`:

```{r}

calcGDP <- function(dat, year = NULL, country = NULL){
  if (!is.null(year)) {
    dat <- dat %>% filter(.data$year == UQ(year) )
  }
  
  if (!is.null(country)) {
   dat <- dat %>% filter(.data$country == UQ(country) )
  }
  
  dat <- dat %>% mutate(gdp = .data$pop * .data$gdpPercap)
        
  return(dat)
}
```

If you've been writing these functions down into a separate R script
(a good idea!), you can load in the functions into our R session by using the
`source` function:

```{r, eval=FALSE}
source("src/functions-lesson.R")
```

OK, so there's a lot going on in this function now. In plain English,
the function now subsets the provided data by year if the year argument isn't
empty, then subsets the result by country if the country argument isn't empty.
Then it calculates the GDP for whatever subset emerges from the previous two steps.
The function then adds the GDP as a new column to the sub-setted data and returns
this as the final result.
You can see that the output is much more informative than a vector of numbers.

Let's take a look at what happens when we specify the year:

```{r}
head(calcGDP(gapminder, year = 1997))
```

Or for a specific country:

```{r}
calcGDP(gapminder, country = "Australia")
```

Or both:

```{r}
calcGDP(gapminder, year = 1997, country = "Australia")
```


> ## Challenge: Finishing our function
> 
> We said that we wanted to be able to handle one _or more_ years and countries
> Modify the function to handle vectors of multiple countries (i.e. `c("Belgium", "Germany")) or years
> 
> > ## Solution
> >
> > ```{r}
> > 
> > calcGDP <- function(dat, year = NULL, country = NULL){
> >   if (!is.null(year)) {
> >    dat <- dat %>% filter(.data$year %in% UQ(year) )
> >   }
> >   
> >  if (!is.null(country)) {
> >    dat <- dat %>% filter(.data$country %in% UQ(country) )
> >  }
> >   
> >  dat <- dat %>% mutate(gdp = .data$pop * .data$gdpPercap)
> >        
> >  return(dat)
> > }
> >```
> >
> {: .solution}
{: .challenge}

> ## Tip: Pass by value
>
> Functions in R almost always make copies of the data to operate on
> inside of a function body. When we modify `dat` inside the function
> we are modifying the copy of the gapminder dataset stored in `dat`,
> not the original variable we gave as the first argument.
>
> This is called "pass-by-value" and it makes writing code much safer:
> you can always be sure that whatever changes you make within the
> body of the function, stay inside the body of the function.
{: .callout}

> ## Tip: Function scope
>
> Another important concept is scoping: any variables (or functions!) you
> create or modify inside the body of a function only exist for the lifetime
> of the function's execution. When we call `calcGDP`, the variables `dat`,
> `gdp` and `new` only exist inside the body of the function. Even if we
> have variables of the same name in our interactive R session, they are
> not modified in any way when executing a function.
{: .callout}


> ## Challenge 3
>
> Test out your GDP function by calculating the GDP for New Zealand in 1987. How
> does this differ from New Zealand's GDP in 1952?
>
> > ## Solution to challenge 3
> >
> > GDP for New Zealand in 1987: 65050008703
> >
> > GDP for New Zealand in 1952: 21058193787
> {: .solution}
{: .challenge}


> ## Challenge 4
>
> The `paste` function can be used to combine text together, e.g:
>
> ```{r}
> best_practice <- c("Write", "programs", "for", "people", "not", "computers")
> paste(best_practice, collapse=" ")
> ```
>
>  Write a function called `fence` that takes two vectors as arguments, called
> `text` and `wrapper`, and prints out the text wrapped with the `wrapper`:
>
> ```{r, eval=FALSE}
> fence(text=best_practice, wrapper="***")
> ```
>
> *Note:* the `paste` function has an argument called `sep`, which specifies the
> separator between text. The default is a space: " ". The default for `paste0`
> is no space "".
>
> > ## Solution to challenge 4
> >
> >  Write a function called `fence` that takes two vectors as arguments, called
> > `text` and `wrapper`, and prints out the text wrapped with the `wrapper`:
> >
> > ```{r}
> > fence <- function(text, wrapper){
> >   text <- c(wrapper, text, wrapper)
> >   result <- paste(text, collapse = " ")
> >   return(result)
> > }
> > best_practice <- c("Write", "programs", "for", "people", "not", "computers")
> > fence(text=best_practice, wrapper="***")
> > ```
> {: .solution}
{: .challenge}

> ## Tip
>
> R has some unique aspects that can be exploited when performing
> more complicated operations. We will not be writing anything that requires
> knowledge of these more advanced concepts. In the future when you are
> comfortable writing functions in R, you can learn more by reading the
> [R Language Manual][man] or this [chapter][] from
> [Advanced R Programming][adv-r] by Hadley Wickham. For context, R uses the
> terminology "environments" instead of frames.
{: .callout}

[man]: http://cran.r-project.org/doc/manuals/r-release/R-lang.html#Environment-objects
[chapter]: http://adv-r.had.co.nz/Environments.html
[adv-r]: http://adv-r.had.co.nz/


> ## Tip: Testing and documenting
>
> It's important to both test functions and document them:
> Documentation helps you, and others, understand what the
> purpose of your function is, and how to use it, and its
> important to make sure that your function actually does
> what you think.
>
> When you first start out, your workflow will probably look a lot
> like this:
>
>  1. Write a function
>  2. Comment parts of the function to document its behaviour
>  3. Load in the source file
>  4. Experiment with it in the console to make sure it behaves
>     as you expect
>  5. Make any necessary bug fixes
>  6. Rinse and repeat.
>
> Formal documentation for functions, written in separate `.Rd`
> files, gets turned into the documentation you see in help
> files. The [roxygen2][] package allows R coders to write documentation alongside
> the function code and then process it into the appropriate `.Rd` files.
> You will want to switch to this more formal method of writing documentation
> when you start writing more complicated R projects.
>
> Formal automated tests can be written using the [testthat][] package.
{: .callout}

[roxygen2]: http://cran.r-project.org/web/packages/roxygen2/vignettes/rd.html
[testthat]: http://r-pkgs.had.co.nz/tests.html
