---
title: "Challenge 3"
author: "David Mawdsley"
params:
  comparator: "United Kingdom"
  year: 2002
output:
  html_notebook: default
---


In this notebook we load and examine the Gapminder data.


```{r setup, include=FALSE}
library(tidyverse)
```

```{r loaddata, echo=FALSE}
# By putting the loading code in a separate chunk, we can still see any 
# messages that may be printed when it loads.  This is useful, since
# messages here could indicate a problem.
gapminder <- read_csv("../data/gapminder-FiveYearData.csv",
                      col_types = cols(
                        country = col_character(),
                        year = col_integer(),
                        pop = col_double(),
                        continent = col_character(),
                        lifeExp = col_double(),
                        gdpPercap = col_double()
                      ))
```

Let's take a look at the data:

```{r, echo=FALSE}
print(gapminder)
```

Life expectancy has, in general, increased over time, although this masks stark falls in some countries:

```{r, echo=FALSE}
ggplot(data = gapminder, aes(x = year, y = lifeExp, colour = continent, group = country)) +
  geom_line()
```


```{r calclongevity, echo=FALSE}

gapminderSorted <- gapminder %>% filter(year == params$year) %>% 
  arrange(desc(lifeExp))

longevityAge <- gapminderSorted$lifeExp[1]
longevityCountry <- gapminderSorted$country[1]

gapminderComparator <- gapminder %>% filter(country == params$comparator, year == params$year)

longevityComparator <- gapminderComparator$lifeExp

```

In `r params$year`, citizens of `r longevityCountry` had the longest average life expectancy in the world, at `r longevityAge`.  By comparison, in `r params$comparator`, the average life expectancy was `r longevityComparator`



