---
title: "Analysis of gapminder data"
author: "David Mawdsley"
output:
  html_notebook
params:
  year: 2002
---


```{r setup, message=FALSE, echo=FALSE}
library(readr)
```

```{r loaddata, echo=FALSE}
gapminder <- read_csv("../data/gapminder-FiveYearData.csv",
                      col_types = cols(
                        country = col_character(),
                        year = col_integer(),
                        pop = col_double(),
                        continent = col_character(),
                        lifeExp = col_double(),
                        gdpPercap = col_double()
                      ))
```

We can look at the mean and median life expectancy for the countries in each contient for 2007:

```{r}
gapminder %>%
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarise(meanexp = mean(lifeExp), medexp = median(lifeExp))
```

## A better approach

What if we wanted to look at a different year?  "2007" appears in two places in the notebook; in the text and in the code. We could easily forget to change one of them if we're interested in a different year. This is where we use parameters, which we define in the header of the document.

We can look at the the mean and median life expectancy for the countries in each continent for `r params$year`:


```{r}
gapminder %>%
  filter(year == params$year) %>% 
  group_by(continent) %>% 
  summarise(meanexp = mean(lifeExp), medexp = median(lifeExp))
```
