---
title: "Data Analysis Using R"
author: "David Mawdsley"
date: "17 October 2017"
output: 
  ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Data Analysis Using R

Initial tasks:
1. Log into a PC (or have your laptop ready to use)
2. Check you can load RStudio
3. Open the course notes page at [https://mawds.github.io/r-novice-gapminder/](https://mawds.github.io/r-novice-gapminder/)
  * There is a link to this slideshow there
4. There is no sign-in sheet.  Attendance is recorded via the [https://goo.gl/forms/YZJ05PzX9tPFEtuV2](feedback form)


## Introduction

* Aims of the day
* Research IT
* Teaching methods


## Aims of the day

* Introduction to R
  + No prior knowledge of R assumed
  + Focus on using R for analysis rather than programming
* Does not teach statistics
* The aim is to teach you enough know how to find out more


## Research IT
Described on the [IT Services website](http://www.itservices.manchester.ac.uk/research/)

Announcements given via the [Research IT blog](https://researchitnews.org/) and newsletter (signup via the blog page)

* [Training courses](http://ri.itservices.manchester.ac.uk/training) teach computing skills for research
* Advice and guidance about research software
* Access to specialist support and consultancy e.g. code reviews
* Access to HPC systems
* [Full list of services on offer](http://www.itservices.manchester.ac.uk/our-services/research/)
Get in touch via the [support centre](http://supportcentre.manchester.ac.uk/)

## Housekeeping

* Fire exit
* Toilets
* Course timing
* Feedback ([form](https://goo.gl/forms/YZJ05PzX9tPFEtuV2) and verbal)

## Why learn R?

* (probably) the most popular language for data-analysis
* It's free
* It's extensible

## 

```{r, echo=FALSE, message=FALSE}
source("https://gist.githubusercontent.com/revodavid/6de6607fbe9a5638922a1aae28e554be/raw/2d137dc73024a1820ba92640b4b08e5a52251d47/CRAN_pkg_history.R")

```

Graph uses code from [http://blog.revolutionanalytics.com/2016/04/cran-package-growth.html](http://blog.revolutionanalytics.com/2016/04/cran-package-growth.html) with updated data.


## Work with maps...
```{r}
library(leaflet)
m = leaflet() %>% addTiles()
m %>% setView( -2.233,53.467, zoom = 16)
```

## Interactive apps (Shiny)

* Perform interactive visualisation and analysis
    + Dashboards
    + Modelling
* [The Shiny user showcase](https://www.rstudio.com/products/shiny/shiny-user-showcase/) has many examples
    + [Travelling salesman example](https://gallery.shinyapps.io/shiny-salesman/)

## Teaching methods

* The course is based on the [Software Carpentry](https://software-carpentry.org) [R for reproducible scientific research](http://swcarpentry.github.io/r-novice-gapminder/) course
    + Abridged
    + Doesn't cover topics that overlap with other Research IT courses
  
## Teaching methods

* The course is _interactive_
    + These are the only slides
* Getting help
    + The Post-its
    + Etherpad

