---
title: "Data Analysis Using R"
author: "David Mawdsley"
date: "17 October 2017"
output: 
  ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

* Aims of the day
* Other courses from Research IT
* Teaching methods


## Aims of the day

* Introduction to R
  + No prior knowledge of R assumed
  + Focus on using R for analysis rather than programming
* Does not teach statistics
* The aim is to teach you enough know how to find out more

## Other courses from Research IT

* We offer courses on [many aspects of scientific computing](http://ri.itservices.manchester.ac.uk/training)
    + All are free for staff and PGRs
* Learning version control is _highly_ recommended
    + [We have a course on this](https://app.manchester.ac.uk/rrgit)

## Why learn R?

* (probably) the most popular language for data-analysis
* It's free
* It's extensible

## 

```{r, echo=FALSE, message=FALSE}
source("https://gist.githubusercontent.com/revodavid/6de6607fbe9a5638922a1aae28e554be/raw/2d137dc73024a1820ba92640b4b08e5a52251d47/CRAN_pkg_history.R")

```

Graph uses code from [http://blog.revolutionanalytics.com/2016/04/cran-package-growth.html](http://blog.revolutionanalytics.com/2016/04/cran-package-growth.html) with updated data.


## Work with maps...
```{r}
library(leaflet)
m = leaflet() %>% addTiles()
m %>% setView( -2.233,53.467, zoom = 16)
```

## Interactive apps (Shiny)

* Perform interactive visualisation and analysis
    + Dashboards
    + Modelling
* [The Shiny user showcase](https://www.rstudio.com/products/shiny/shiny-user-showcase/) has many examples
    + [Travelling salesman example](https://gallery.shinyapps.io/shiny-salesman/)

## Teaching methods

* The course is based on the [Software Carpentry](https://software-carpentry.org) [R for reproducible scientific research](http://swcarpentry.github.io/r-novice-gapminder/) course
    + Abridged
    + Doesn't cover topics that overlap with other Research IT courses
  
## Teaching methods

* The course is _interactive_
    + These are the only slides
* Getting help
    + The Post-its
    + Etherpad

